Creating an Arrow IPC file from the NYC 311 data

### 2023-06-18 nyc311clean.qmd
Clean up some of the explanations.

We show conversion of a CSV file containing data on 311 service calls during March, 2023 in New York City to the [Arrow](https://arrow.apache.org) IPC (Inter-Process Communications) file format (also called the Feather V2 format) which can be read quickly and easily into many different data science environments.
This conversion is done in [Julia](https://julialang.org) but it could equally well have been done in a number of different languages, whichever is most convenient for the person doing the conversion.

@@

```{julia}
#| panel: fill
open("./data/311_half_Mar_2023.csv", "r") do io
    println(readline(io))
    println(readline(io))
    seekstart(io)
end;
```

::: {.callout-note collapse="false"}
### Need to scroll right to see all of the output lines

@@

Note that the character on the right of the title bar for callout blocks like this is a toggle to show or hide the contents of the block.
:::

@@

```{julia}
df = CSV.read(
    "./data/311_half_Mar_2023.csv",  # file name
    DataFrame;                       # output table type
    normalizenames=true, # convert column names to valid symbols
    missingstring=[      # strings that are used to indicate missing vals
        "",
        "Unspecified",
        "unspecified",
        "Unknown",
        "unknown",
        "NA",
    ],
    dateformat=dateformat"m/d/Y I:M:S p", # DateTimes look like this
    downcast=true,       # use smallest type of Int that can represent the column 
    stripwhitespace=true # strip leading and trailing whitespace 
)
```

:::{.callout-note collapse="true"}
### normalize names
Convert the column names in the first row of the CSV file into a form suitable as a `Symbol` in `Julia`.
In particular, strip whitespace at the beginning and end of the name and replace embedded blanks or `.` characters by underscores.
:::

@@

:::{.callout-note collapse="true"}
### downcast
For columns of integer values, choose the smallest type that can represent all the values in the column.
:::

@@

In the output (you must scroll horizontally to see it all) under the column name is an indicator of the type.
A type name that ends in `?` indicates that missing values are present in the column.
Thus the `Unique_Key` is stored as 32-bit signed integers with no missing values.
The `Created_Date` is a `DateTime` with no missing values but the `Closed_Date`, which is also a `DateTime`, does have missing values.

Some columns are described as `String` whereas others are described as `String7` or `String15`.
These are storage-saving representations used when the whole column consists of short character strings.
There is an anomaly in these short strings for the `Agency` column.
The `Agency` acronyms should be 4 or 5 characters at most but at least one (it happens to be `NYC311-PRD`) must be greater than 7 characters to force `String15` storage.

```{julia}
show(unique(df.Agency))
```

The `Agency` and `Agency_Name` columns should be in one-to-one correspondence but they are not.

@@

@@

The other problem shown in the types is that the values in the `Incident_Zip` column are stored as strings when they should be integers or missing.

```{julia}
levels(df.Incident_Zip)
```

and, of course, the problem is that someone got creative and used `na` as a missing value indicator.

@@

```{julia}
df = CSV.read(
    "./data/311_half_Mar_2023.csv",  # file name
    DataFrame;                       # output table type
    normalizenames=true, # convert column names to valid symbols
    missingstring=[      # strings that are used to indicate missing vals
        "",
        "Unspecified",
        "unspecified",
        "Unknown",
        "unknown",
        "NA",
        "na",
    ],
    dateformat=dateformat"m/d/Y I:M:S p", # DateTimes look like this
    downcast=true,       # use smallest type of Int that can represent the column 
    stripwhitespace=true # strip leading and trailing whitespace 
)
```

@@

### Are the `Created_Date` values in the specified range?

@@

```{julia}
earlyclosing = (!ismissing).(df.Closed_Date) .&& (df.Closed_Date .< df.Created_Date)
immediateclosing = (!ismissing).(df.Closed_Date) .&& (df.Closed_Date .== df.Created_Date)
(early = count(earlyclosing), immediate = count(immediateclosing))
```

:::{.callout-note collapse="true"}
### What are all those dots for?
Julia provides "syntactic sugar" to vectorize a scalar operation with a `.` following a function name (e.g. `(!ismissing).`) or preceding an operator (e.g. `.<`)
:::

@@

:::{.callout-note collapse="true"}
### Use of `:` as an index indicates all values for that index
In this case the use of `:` as the second index indicates that all the columns should be included in the subset.
:::

@@

This is a tricky question because the values of `Latitude` and `Longitude` are stored to 17 significant digits, which is much more accuracy than could possibly be measured.
To check for consistency we need to convert the `Location` to a `Latitude-Longitude` pair then compare these values to the recorded `Latitude` and `Longitude` by employing a tolerance in the comparison of floating-point values.
We'll leave that as an exercise for the reader and just assure you that the `Location` column is redundant and can be dropped.

@@

```{julia}
#| output: false
fn = Arrow.write(
    "./data/311_half_Mar_2023.arrow",
    select(df, Not(:Location)),
)
```

@@

The columns that are described as `<fct>` in the `glimpse` output from R are characterized as `dictionary` in the output from `fea.read_table` and listed as `cat` (for "categorical") in the Polars data frame.
The conversion of a column of character strings, many of which are repetitions, to a combination of a vector of the unique values and a small-sized integer index into that vector, can result in considerable storage saving, at the expense of a trivial amount of processing.
For example, storing the `Agency_Name` as a vector of strings takes about 34 MB but, after dictionary encoding, it takes up about 1/8 MB.

>>>>

We show conversion of a CSV file containing data on 311 service calls in New York City during March, 2023 to the [Arrow](https://arrow.apache.org) IPC (Inter-Process Communication) file format (also called the Feather V2 format), which can be read quickly and easily into several different data science environments.
This conversion is done in [Julia](https://julialang.org) but it could equally well have been done in a different language, if that is more convenient for the person doing the conversion.

@@

```{julia}
#| panel: fill
open("./data/311_half_Mar_2023.csv", "r") do io # open the file
    println(readline(io))     # read and print the first line
    println(readline(io))     # read and print the next line
end;  # this open/do/end construction automatically closes the file when done
```

::: {.callout-note collapse="false"}
### Need to scroll right to see the whole output line

@@

Note that the character on the right of the title bar for callout blocks like this, currently a `v`-shape for this block, is a toggle to show or collapse the contents of the block.
Try clicking on that character to collapse the contents of the block.

In what follows the callout blocks are initially collapsed and you must click on the `>` character on the right end of the title bar to show the contents.
:::

@@

```{julia}
df = CSV.read(
    "./data/311_half_Mar_2023.csv",  # file name
    DataFrame,           # output table type
    normalizenames=true, # convert column names to valid symbols
    missingstring=[      # strings that are used to indicate missing values
        "",
        "Unspecified",
        "unspecified",
        "Unknown",
        "unknown",
        "NA",
    ],
    dateformat=dateformat"m/d/Y I:M:S p", # DateTimes look like this
    downcast=true,       # use smallest type of Int that can represent the column 
    stripwhitespace=true # strip leading and trailing whitespace 
)
```

:::{.callout-note collapse="true"}
### normalizenames
Convert the column names from the CSV file into a form that can be parsed as a `Symbol` in `Julia`.
In particular, strip whitespace at the beginning and end of the name and replace embedded blanks or `.` characters by underscores.

Columns of a DataFrame (and other type of tables) are usually extracted as "properties" of the table, either as, e.g. `getproperty(df, :Unique_Key)`, or, more commonly, with the dot operator as, e.g. `df.Unique_Key`.
The expression `:Unique_Key` or the right operand of `df.Unique_Key` evaluates to a Symbol.
If the column name has an embedded blank or `.` this form cannot be used and the name must be quoted, either as `Symbol("Unique.Key")` or `var"Unique.Key"`.

Thus it is an advantage to rewrite the column names so that each name can be used "as is" with the dot operator.
:::

@@

:::{.callout-note collapse="true"}
### downcast
For columns of integer values, choose the smallest integer type, such as `Int8`, `Int16`, etc., that can represent all the values in the column.
:::

@@

In the output (remember, you must scroll horizontally to see it all) there is a description of the type of the elements of the column directly under the column name.
A type name that ends in `?` indicates that missing values are present in the column.
Thus, for example, the `Unique_Key` is designated `Int32` meaning the values are stored as 32-bit signed integers with no missing values.
The `Created_Date` column, designated `DateTime`, is a `DateTime` column with no missing values but the `Closed_Date`, designated `DateTime?`, is also a `DateTime`, but can (and does) have missing values.

Some columns are designated as `String` whereas others are designated as `String7` or `String15`.
These are storage-saving representations used when the whole column consists of short character strings.

There is an anomaly in these short strings for the `Agency` column.
The `Agency` acronyms should be 5 characters at most, which would be stored as `String7`, but they are stored as `String15`.

To see why, we check the unique combinations of the `Agency` and `Agency_Name` columns

@@

These columns should be in one-to-one correspondence but they are not.

@@

The other problem shown in the types is that the values in the `Incident_Zip` column are stored as `String7?` when they should be `Int32?`.

Checking the unique values

```{julia}
sort(unique(df.Incident_Zip))
```

shows the problem is that someone got creative and used `na` as a missing value indicator.

@@

```{julia}
df = CSV.read(
    "./data/311_half_Mar_2023.csv",  # file name
    DataFrame,           # output table type
    normalizenames=true, # convert column names to valid symbols
    missingstring=[      # strings that are used to indicate missing vals
        "",
        "Unspecified",
        "unspecified",
        "Unknown",
        "unknown",
        "NA",
        "na",
    ],
    dateformat=dateformat"m/d/Y I:M:S p", # DateTimes look like this
    downcast=true,       # use smallest type of Int that can represent the column 
    stripwhitespace=true # strip leading and trailing whitespace 
)
```

@@

### Are the `Created_Date` values all from March, 2023?

@@

```{julia}
earlyclosing = @. !ismissing(df.Closed_Date) && (df.Closed_Date < df.Created_Date)
immediateclosing = @. !ismissing(df.Closed_Date) && (df.Closed_Date == df.Created_Date)
(early = count(earlyclosing), immediate = count(immediateclosing))
```

:::{.callout-note collapse="true"}
### What does `@.` do?

[Dot vectorization](https://docs.julialang.org/en/v1/manual/functions/#man-vectorized) in Julia provides "syntactic sugar" to vectorize a scalar operation.
Prefacing an expression with `@.` applies dot vectorization to all the function calls and operators in the expression. 
:::

@@

:::{.callout-note collapse="true"}
### Use of `:` as an index indicates all values for that axis
In this case the use of `:` as the second index indicates that all the columns should be included in the subset.
:::

@@

This is a tricky question because the values of `Latitude` and `Longitude` are stored to 17 significant digits, which is much more accuracy than could possibly be measured, but equality comparisons, either as floating-point numbers or as strings with 17 significant digits, would require an exact match.

To check for consistency we need to convert the `Location` to a `Latitude-Longitude` pair of floating point numbers then check for approximate equality of these values to the recorded `Latitude` and `Longitude`.
We'll leave that as an exercise for the reader and just assure you that the `Location` column is redundant and can be dropped.

@@

```{julia}
#| output: false
fn = Arrow.write(    # returns the file name
    "./data/311_half_Mar_2023.arrow",
    select(df, Not(:Location)),
)
```

@@

The columns that are described as `<fct>` in the `glimpse` output from R (and `dictionary` in the output from `fea.read_table` and `cat` (for "categorical") in the Polars data frame) are `factor` representations, consisting of a (short) vector of the unique values, called the `levels` of the factor, and an integer vector of indices into the levels.
The conversion of a column of character strings, many of which are repetitions, to a
factor or "dictionary encoded" representation, can result in considerable storage saving, at the expense of a trivial amount of processing.
For example, storing the `Agency_Name` as a vector of strings takes about 34 MB but, after dictionary encoding, it takes up about 1/8 MB.

The output from `fea.read_table` is often the most informative representation of the structure of the table.
